---
title: 'the osmdata Package '
author: 'Jonathan Jupke '
date: '2021-03-01'
slug: the-osmdata-package
categories:
  - GIS
tags:
  - GIS
toc: no
images: ~
---

Here I will introduce you to the  **osmdata** package. 
osm is short for open street map and you can use the package to access the data from osm directly from R. 

```{r eval=F}
install.packages("osmdata")
libary(osmdata)
library(sf)
library(dplyr)
library(tmap)
library(magrittr)
```
```{r echo=F, message = F, results = 'hide'}
library(osmdata)
library(sf)
library(dplyr)
library(tmap)
library(magrittr)
```

osm uses the read-only overpass API to download data from Open Street Map and read them in the formats of
**sp**, **sf** or **silicate**. 
A complete call with osmdata looks like his: 

```{r eval = F}
x <-
  getbb("Landau") %>% 
        opq() %>%
        add_osm_feature(key = 'name',
                        value = "Queich") %>%
        osmdata_sf()
```

As you can see, four different functions are part of the call. 

1. `getbb()`
2. `opq()`   
3. `add_osm_feature()`
4. `osmdata_sf()`

Lets go through them one after another to see what they do. 
`getbb()` returns the bounding box of some place in the world. 
The only argument you usually have to supply is `place_name`.

```{r}
ld = getbb("Landau")
```

This worked, and returned the bounding box of Landau in longitude and latitude. 
Using our new bounding box we can now call `opq()`. 
Of course you can also write a custom bbox directly into `opq()`.
This function builds a query to Overpass (**o**ver**p**ass **q**uery).

```{r}
ld = opq(bbox = ld)
ld
```

This is no spatial object yet but just the request (query) that will be send to the API. 
This request is further expanded in `add_osm_feature()`. 
All items in the Open Street Map data have a **key** and a **value**. 
A few example are key: amenity - value: bar, cafe or fast_food; key: building - value: school, public or church, key: landuse - value: basin, cementery, grass. 
See [here](https://wiki.openstreetmap.org/wiki/Map_Features) for a complete overview of all keys and features. 
Let's download one of each. 

```{r}
ld_amenity = add_osm_feature(ld, key = "amenity", value = "bar") 
ld_building = add_osm_feature(ld, key = "building", value = "school") 
ld_landuse = add_osm_feature(ld, key = "landuse", value = "grass") 
```

Now our queries are complete and we can download the data using `osmdata_sf()`. 

```{r results="hide"}
ld_amenity  %<>% osmdata_sf() 
ld_building %<>% osmdata_sf() 
ld_landuse  %<>% osmdata_sf() 
```

The suffix of `osmdata_sf()` specifies that we want the downloaded data to be in the `sf` format. 
Alternatives are: `osmdata_sp`,`osmdata_sc` and `osmdata_xml`.

The resulting objects are lists which have the different geometries as elements.
Below we plot the three data sets together.

```{r}
tmap_mode("view")
tm_shape(ld_amenity$osm_points) + tm_dots(col = "red") + 
  tm_shape(ld_building$osm_polygons) + tm_polygons(col = "blue") + 
  tm_shape(ld_landuse$osm_polygons) + tm_polygons(col = "green")
```

