---
title: 'Analyzing the antTraits data with CQO and CAO '
author: Jonathan Jupke
date: '2020-12-31'
slug: analyzing-the-anttraits-data-with-cqo-and-cao
categories: []
tags:
  - CQO
  - CAO
  - R
  - antTraits
  - MOD3
comments: no
images: ~
---


# TO DO
add references to Yee 2015




load libraries and custom made cqo_residual_plot function (hope to put this into misc package at some point)
```{r}
pacman::p_load(data.table, dplyr, ggplot2, purrr, VGAM, mvabund)
source("../../../../../01_Uni/01_Lehre/05_MOD3-Lecture/002_r_scripts/cqo/cqo_residual_plot.R")
```

```{r echo = F}
dir_cao = "../../../../../01_Uni/01_Lehre/05_MOD3-Lecture/003_processed_data/cao/antTraits/"
dir_cqo = "../../../../../01_Uni/01_Lehre/05_MOD3-Lecture/003_processed_data/cqo/antTraits/"
cao_p = readRDS(file.path(dir_cao, "cao.RDS"))
```


load data 
```{r}
data("antTraits")

bio = antTraits$abund
env = antTraits$env
env = scale(env)
```

In Yee (2015, p.241), Thomas Yee suggests to keep S <= 10, n <=500 and p <=5. We will stick to this. 
We have 41 species so I pick the 10 most abundant ones.

```{r}
colSums(bio) %>% 
        sort(decreasing=T) %>% 
        names %>% 
        .[11:10]
comb = cbind(bio,env)
```

We start with a CAO to evaluate unimodality. 

A non-linear degree of freedom (df1.nl) of 1.5 approximates CQO. By setting df1.nl to 2 we apply a liberal inclusion criterion. I set a seed to ensure replicability. 

```{r eval = F}
set.seed = 1
cao_p = cao(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Pheidole.sp..A,
    Rhytidoponera.metallica.sp..A,
    Pheidole.sp..E,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.leae,
    Iridomyrmex.mjobergi,
    Heteroponera.sp..A,
    Nylanderia.sp..A
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = poissonff,
  data = comb,
  Rank = 1,
  df1.nl = 2,
  Bestof = 30 
)
```

With the second line below we evaluate if a global minimum was reached. If the number that is returned there is below 3 we should rerun the CAO and possibly increase the *Bestof* argument. Here 8 from 30 models have the same deviance which means that we are likley to have found a global minimum. 

```{r}
cao_p_summary = summary(cao_p)
sum(
  round(cao_p_summary@misc$deviance.Bestof,0) 
  == 
  round(min(cao_p_summary@misc$deviance.Bestof), 0)
  )
par(mfrow=c(3,4))
plot(
  cao_p,
  lcol = "blue",
  lwd = 2,
  ylim = c(-5, 5),
  xlab = "",
  ylab = ""
)
```

Most species do not respond unimodally so we replace them with new ones. 

```{r eval = F}
cao_p = cao(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.rothsteini,
    Notoncus.ectatommoides,
    Tapinoma.sp..A,
    Rhytidoponera.sp..B,
    Iridomyrmex.suchieri,
    Melophorus.sp..F,
    Aphaenogaster.longiceps
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = poissonff,
  data = comb,
  Rank = 1,
  df1.nl = 2,
  Bestof = 30
)
```

```{r echo = F}
cao_p = readRDS(file.path(dir_cao, "cao2.RDS"))
```


```{r}
cao_p_summary = summary(cao_p)
sum(round(cao_p_summary@misc$deviance.Bestof,0) == round(min(cao_p_summary@misc$deviance.Bestof), 0))
par(mfrow=c(3,4))
plot(cao_p, lcol = "blue", lwd = 2, ylim = c(-5, 5), xlab = "", ylab = "")
```

Some species still look bad (e.g. ). Let's go through the motions one more time. 
```{r eval = F}
cao_p = cao(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.rothsteini,
    Iridomyrmex.suchieri,
    Aphaenogaster.longiceps,
    Iridomyrmex.purpureus,
    Camponotus.claripes,
    Tetramorium.sp..A,
    Meranoplus.sp..A,
    Monomorium.sydneyense
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = poissonff,
  data = comb,
  Rank = 1,
  df1.nl = 2,
  Bestof = 30
)
```

```{r echo = F}
cao_p = readRDS(file.path(dir_cao, "cao3.RDS"))
```

```{r}
cao_p_summary = summary(cao_p)
sum(round(cao_p_summary@misc$deviance.Bestof,0) == round(min(cao_p_summary@misc$deviance.Bestof), 0))
par(mfrow=c(3,4))
plot(cao_p, lcol = "blue", lwd = 2, ylim = c(-5, 5), xlab = "", ylab = "")
```


Now we can fit a CQO with Poisson response and one latent variable. This first model estimates tolerances separately for each species (*eq.toler* = FALSE) 
```{r eval = F}
cqo_p_ut = cqo(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.rothsteini,
    Aphaenogaster.longiceps,
    Iridomyrmex.purpureus,
    Tetramorium.sp..A,
    Meranoplus.sp..A
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = poissonff,
  data = comb,
  Rank = 1,
  Bestof = 20,
  eq.toler = F
)
```

```{r echo=F}
cqo_p_ut = readRDS(file.path(dir_cqo, "cqo_p_ut.RDS"))
```

```{r}
cqo_p_summary = summary(cqo_p_ut)
sum(round(cqo_p_summary@misc$deviance.Bestof,0) == round(min(cqo_p_summary@misc$deviance.Bestof), 0))
Tol(cqo_p_ut)[1, 1, ]
is.bell(cqo_p_ut)
```

```{r eval = F}
cqo2_p_ut = cqo(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.rothsteini,
    Aphaenogaster.longiceps,
    Iridomyrmex.purpureus,
    Tetramorium.sp..A,
    Meranoplus.sp..A
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = poissonff,
  data = comb,
  Rank = 2,
  Bestof = 30,
  eq.toler = F
)
```

```{r echo=F}
cqo2_p_ut = readRDS(file.path(dir_cqo, "cqo2_p_ut.RDS"))
```

```{r}
cqo2_p_ut_summary = summary(cqo2_p_ut)
sum(round(cqo2_p_ut_summary@misc$deviance.Bestof,0) == round(min(cqo2_p_ut_summary@misc$deviance.Bestof), 0))
Tol(cqo2_p_ut)
is.bell(cqo2_p_ut)
AIC(cqo_p_ut)
AIC(cqo2_p_ut)
```

```{r}
cqo_resid_plot(list(cqo_p_ut,
                    cqo2_p_ut))
```

```{r eval =F}
cqo_nb_ut = cqo(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.rothsteini,
    Aphaenogaster.longiceps,
    Iridomyrmex.purpureus,
    Tetramorium.sp..A,
    Meranoplus.sp..A
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = negbinomial,
  data = comb,
  Rank = 1,
  Bestof = 40,
  eq.toler = F
)
```

```{r echo=F}
cqo_nb_ut = readRDS(file.path(dir_cqo, "cqo_nb_ut.RDS"))
```

```{r warning=F}
cqo_nb_ut_summary = summary(cqo_nb_ut)
sum(round(cqo_nb_ut_summary@misc$deviance.Bestof,0) == round(min(cqo_nb_ut_summary@misc$deviance.Bestof), 0))
Tol(cqo_nb_ut_summary)
is.bell(cqo_nb_ut_summary)
cqo_resid_plot(list(cqo_nb_ut, cqo_p_ut))
# lets look at the plots 
par(mfrow=c(1,1))
lvplot(cqo_p_ut, label = T)
#persp(cqo_p_ut, label = T)
lvplot(cqo2_p_ut, label = T)
#persp(cqo2_p_ut, label = T)

```

```{r eval=F}
cqo2_p_et = cqo(
  formula = cbind(
    Iridomyrmex.rufoniger,
    Iridomyrmex.bicknelli,
    Camponotus.consobrinus,
    Monomorium.rothsteini,
    Aphaenogaster.longiceps,
    Iridomyrmex.purpureus,
    Tetramorium.sp..A,
    Meranoplus.sp..A
  ) ~
    Bare.ground + Canopy.cover + Shrub.cover + Volume.lying.CWD + Feral.mammal.dung,
  family = poissonff,
  data = comb,
  Rank = 2,
  Bestof = 50,
  eq.toler = T
)
```

```{r}
cqo2_p_et = readRDS(file.path(dir_cqo, "cqo2_p_et.RDS"))
```

```{r warning=F}
cqo2_p_et_summary = summary(cqo2_p_et)
sum(round(cqo2_p_et_summary@misc$deviance.Bestof,0) == round(min(cqo2_p_et_summary@misc$deviance.Bestof), 0))

# trivial 
is.bell(cqo2_p_et)

cqo_resid_plot(list(cqo2_p_et, cqo2_p_ut))
lvplot(cqo2_p_et, label = T)
persp(cqo2_p_et, label = T)

concoef(cqo2_p_et)
```

