---
title: 'The rayshader R package '
author: 'Jonathan Jupke '
date: '2021-02-10'
slug: the-rayshader-r-package
categories:
  - GIS
tags:
  - GIS
  - R
  - R package
  - rayshader
  - raster
comments: no
images: ~
---

In this entry we will go through the basics of **rayshader**. 
Rayshader is one of those packages that you see again and again when you follow people from the R-spatial community on twitter. 
Especially the package's creator Tyler Morgan-Wall (see [here](https://www.tylermw.com/) for his website and [here](https://twitter.com/tylermorganwall) for his Twitter) posts videos and images of things he did with it on an almost daily basis. 
And he has all the reasons to do so.
I don't know any other package that enables you to create such stunning 3D graphics with R. 
If you have never seen this package before I am quite sure you are going to be surprised by what is possible. 

# Setup

First things first, we need to install and load the **rayshader** package, as well as some other packages we will need along the way. 

```{r eval=F}
install.packages("rayshader")
library(rayshader)
library(raster)
library(dplyr)
```
```{r echo=F}
library(rayshader)
library(raster)
library(dplyr)
```

We will use an example raster file from Tyler Morgan-Walls website. 
We can  download the files with the following code: 

```{r eval=F}
loadzip = tempfile()
download.file("https://tylermw.com/data/dem_01.tif.zip", loadzip)
localtif = raster(unzip(loadzip, "dem_01.tif"))
unlink(loadzip)
```
```{r echo=F}
localtif = raster("~/04_Blog/blog/data/rayshader/dem_01.tif")
```

Before we do anything fancy let's have look at our new raster. 
It is a DEM with 505 rows and 550 columns and each cell has size of 33.3m * 33.3m.
There are strong differences in height here, from 0 to 971m. 
```{r}
localtif
plot(localtif)
```

To work with *rayshader** we will need to transform this to a matrix with the `raster_to_matrix()` function. 
Then we can start with add a specific [texture]{style="color:CornflowerBlue"} as well as shadows.  
```{r}
elmat = raster_to_matrix(localtif)
elmat %>%
  sphere_shade(texture = "desert") %>%
  plot_map()
```
The shadows are computed for a default sun angle, but we can change that angle if we like. 
```{r}
elmat %>%
  sphere_shade(sunangle = 45, texture = "desert") %>%
  plot_map()
```

Now you might have noticed that the areas in the valley look suspiciously flat. 
That's because they are water surfaces. 
With the two functions `detect_water()` and `add_water()` we can add water to our map. 
`detect_water()` returns a binary matrix with the same dimensions as `elmat`. 
A one indicates that the corresponding cell in `elmat` contains water and a zero that the cell does not. 
With the [cutoff]{style="color:CornflowerBlue"} argument you can specify how high the water should be, i.e. which cells are classified as carrying water. 
High values mean lower water level. 

```{r}
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat, cutoff = 0.9), color = "desert") %>%
  plot_map()
```

Now we don't have any shadows on the water yet. 
But that can be done with `add_shadow()`.


```{r}
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat), color = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>%
  plot_map()
```

In addition to rayshade we can also ad ambient shade, to make it look even better. 

```{r}
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat), color = "desert") %>%
  add_shadow(ray_shade(elmat), 0.5) %>%
  add_shadow(ambient_shade(elmat), 0) %>%
  plot_map()
```

All this was only the preparation though for the 3D capabilities of **rayshader**.

```{r}
elmat %>%
  sphere_shade(texture = "desert") %>%
  add_water(detect_water(elmat), color = "desert") %>%
  add_shadow(ray_shade(elmat, zscale = 3), 0.5) %>%
  add_shadow(ambient_shade(elmat), 0) %>%
  plot_3d(elmat, zscale = 10, fov = 0, theta = 135, zoom = 0.75, phi = 45, windowsize = c(1000, 800))
Sys.sleep(0.2)
render_snapshot()