---
title: 'R package of the week: xray '
author: Jonathan Jupke
date: '2021-01-25'
slug: r-package-of-the-week-xray
categories: []
tags:
  - R package
  - xray
  - antTraits
comments: no
images: ~
---

For this first post in the series, we will look at small but nice packge called x-ray.
Just like a doctor can use x-rays whether something is wrong with your 

In this post I want to show you, how you can use the xray R package to ...
As an example data set we will use the AntTraits data set from the mvabund package. 
```{r}
pacman::p_load(mvabund, xray)
data("antTraits")
data1 = antTraits$abund
data2 = antTraits$env
```

The first function,`anomalies()`, is great to get an initial impression of your data. 
You only have to provide with one argument (a data set) and it returns two objects: 
1. *<span style="color:CornflowerBlue">variables</span>*
2. *<span style="color:CornflowerBlue">problem_variables</span>*
The function evaluates the following properties: number of NAs, number of zeros, number of blank or empty cells, number of infinite entries, the number of distinct values, the variable class and finally the number of cells that are considered **anomalous**. 
The variables to two prefixes: q (quantity) and p (percent). 
We can supply the function with two additional arguments; 
*<span style="color:CornflowerBlue">anomaly_threshold</span>* and *<span style="color:CornflowerBlue">distinct_threshold</span>* which determine the maximum number of anomalous observations a variables can have and the minimum number of distinct values it needs to have, before it is considered problematic. 
```{r warning=FALSE}
data1_anom = anomalies(data1)
data1_anom$variables
```

In our case, we don't have any blanks or NAs but lots of zeros which is common or abundance data. 
The next function comes in handy to determine the distributions you variables follow. 
```{r}
distributions(data1[,1:3])
distributions(data2)
```


